---
title: "ASCEND to the MAX - how to get the most out of the ASCEND package"
author: "Anne Senabouth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASCEND to the MAX - how to get the most out of the ASCEND package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Before you begin
### System Requirements
Datasets produced by scRNASeq experiments are very large, ranging from a few hundred to a million cells. The number of cells affect the amount of computational resources required to process the dataset – therefore, you need to determine if you have enough computational power and time to complete the analysis. ASCEND can comfortably analyse datasets of up to 10,000 cells on a single machine with 8GB of RAM and a quad-core CPU. Larger datasets should be run on a High Performance Cluster (HPC).

### Configuring BiocParallel
This package makes extensive use of [BiocParallel](http://bioconductor.org/packages/release/bioc/html/BiocParallel.html), enabling ASCEND to make the most of your computer's hardware. As each system is different, BiocParallel needs to be configured by the user. Here are some example configurations.

#### Unix/Linux/MacOS (Single Machine)
```{r SetupNix, eval = TRUE}
library(BiocParallel)
ncores <- parallel::detectCores() - 1
register(MulticoreParam(workers = ncores, progressbar=TRUE), default = TRUE)

```

#### Windows (Single Machine - Quad-core system)
```{r SetupWin, eval = FALSE}
library(BiocParallel)
workers <- 3 # Number of cores on your machine - 1
register(SnowParam(workers = workers, type = "SOCK", progressbar = TRUE), default = TRUE)
```

#### Loading ASCEND
Please use `devtools` to load the development version of `ASCEND`.

```{r LoadASCEND}
devtools::load_all("/Users/a.senabouth/CodeRepositories/ASCEND")
```

You can also use devtools' `install_github` function to install the package, and then load it as normal.

```{r LoadASCENDGit, eval = FALSE}
devtools::install_github("IMB-Computational-Genomics-Lab/ASCEND")
library(ASCEND)
```

