---
title: "ASCEND to the MAX - how to get the most out of the ASCEND package"
author: "Anne Senabouth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ASCEND to the MAX - how to get the most out of the ASCEND package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Before you begin
### System Requirements
Datasets produced by scRNASeq experiments are very large, ranging from a few hundred to a million cells. The number of cells affect the amount of computational resources required to process the dataset – therefore, you need to determine if you have enough computational power and time to complete the analysis. ASCEND can comfortably analyse datasets of up to 10,000 cells on a single machine with 8GB of RAM and a quad-core CPU. Larger datasets should be run on a High Performance Cluster (HPC).

We have tested this package with datasets ranging from 100 to 20,000 cells. Generally, increasing the number of cpus will decrease the processing time of functions, while larger datasets require more RAM. 

### Configuring BiocParallel
This package makes extensive use of [BiocParallel](http://bioconductor.org/packages/release/bioc/html/BiocParallel.html), enabling ASCEND to make the most of your computer's hardware. As each system is different, BiocParallel needs to be configured by the user. Here are some example configurations.

#### Unix/Linux/MacOS (Single Machine)
```{r SetupNix, eval = TRUE}
library(BiocParallel)
ncores <- parallel::detectCores() - 1
register(MulticoreParam(workers = ncores, progressbar=TRUE), default = TRUE)

```

#### Windows (Single Machine - Quad-core system)
```{r SetupWin, eval = FALSE}
library(BiocParallel)
workers <- 3 # Number of cores on your machine - 1
register(SnowParam(workers = workers, type = "SOCK", progressbar = TRUE), default = TRUE)
```

#### Loading ASCEND
Please use `devtools` to load the development version of `ASCEND`.

```{r LoadASCEND}
devtools::load_all("/Users/a.senabouth/CodeRepositories/ASCEND")
```

You can also use devtools' `install_github` function to install the package, and then load it as normal.

```{r LoadASCENDGit, eval = FALSE}
devtools::install_github("IMB-Computational-Genomics-Lab/ASCEND")
library(ASCEND)
```

#### Required packages
ASCEND requires the following packages:
##### CRAN
* [Matrix](https://cran.r-project.org/web/packages/Matrix)
* [ggplot2](https://cran.r-project.org/web/packages/ggplot2)
* [data.table](https://cran.r-project.org/web/packages/data.table)
* [dplyr](https://cran.r-project.org/web/packages/dplyr)
* [reshape2](https://cran.r-project.org/web/packages/reshape2)
* [Rtsne](https://cran.r-project.org/web/packages/Rtsne)
* [limSolve](https://cran.r-project.org/web/packages/limSolve)
* [ggbeeswarm](https://cran.r-project.org/web/packages/ggbeeswarm)
* [dynamicTreeCut](https://cran.r-project.org/package=dynamicTreeCut)
* [dendextend](https://cran.r-project.org/web/packages/dendextend)
* RColorBrewer(https://cran.r-project.org/web/packages/RColorBrewer)

##### Bioconductor
* [Biobase](http://bioconductor.org/packages/release/bioc/html/Biobase.html)
* [BiocParallel](http://bioconductor.org/packages/release/bioc/html/BiocParallel.html)
* [scater](http://bioconductor.org/packages/release/bioc/html/scater.html)
* [scran](http://bioconductor.org/packages/release/bioc/html/scran.html)
* [DESeq](http://bioconductor.org/packages/release/bioc/html/DESeq.html)
