---
title: "Visualising data with ASCEND"
author: "Anne Senabouth"
date: "26/09/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Visualising data with ASCEND
`ASCEND` provides functions for generating PCA, TSNE and MDS plots that can be supplied with a condition list. The conditions can be of clustering results, batch information or the expression of specific genes.

#### Generating a condition list
The function `MakeGeneCondition` generates a condition list, labelling if a cell is expressing the selected gene, or not. This can be used to highlight cells that are expressing genes on plots generated by the `ASCEND` package.

We will label cells that are expressing the IGLC2 transcript.

```{r ConditionList}
condition.list <- MakeGeneCondition(aem.set, "BRN3B")
```

#### PlotPCA
```{r PlotPCA, fig.show="hold", fig.align="center", fig.width = 5}
pca.plot <- PlotPCA(aem.set, condition.list = condition.list)
print(pca.plot)
```

#### PlotTSNE
```{r PlotTSNE, fig.show="hold", fig.align="center", fig.width = 5}
tsne.plot <- PlotTSNE(aem.set, PCA = TRUE, condition.list = condition.list)
print(tsne.plot)
```

#### PlotMDS
```{r PlotMDS, fig.show="hold", fig.align="center", fig.width = 5}
mds.plot <- PlotMDS(aem.set, PCA = FALSE, condition.list = condition.list)
print(mds.plot)
```
